---
title: "Capstone Analysis"
author: "Jishnu Sathees Lalu"
date: "today"
format: html
---

## Load packages

```{r}
library(readxl)
library(dplyr)
library(writexl)
library(ggplot2)
library(tidyr)
```

## Import the data

The data are read from the Excel file

```{r}
tb <- read_excel("~/Desktop/Research/R classes/Capstone Project/clinical parameters_TBpatients.xlsx")
```

##Bring the date of assessment column first and rename it to "Start"

```{r}
tb <- tb |> rename(start=`Date of Assessment`)
tb <- tb |> relocate(start,.before = end)
```

##create a unique ID from row ID
```{r}
tb <- tb |> mutate(id=row_number(),.before = 1)
```

##find mean age. Replace any missing value with mean
```{r}
mean(tb$Age)
sum(is.na(tb$Age))
which(is.na(tb$Age))
tb <- tb |> mutate(Age=ifelse(is.na(Age), mean(Age, na.rm=TRUE), Age))
round(mean(tb$Age),2)
t.test(tb$Age)$conf.int
round(t.test(tb$Age)$conf.int,2)
```
Mean age of the study population was found to be 32.27 years with 95% CI (30.99 - 33.56 years)

##find the mean delay in assessment. Replace any missing value
```{r}
sum(is.na(tb$start))
sum(is.na(tb$end))
tb <- tb |> mutate(delay=as.numeric(as.Date(end)-as.Date(start)))
mean(tb$delay, na.rm=TRUE)
round(mean(tb$delay),2)
round(t.test(tb$delay)$conf.int,2)
```
The mean delay in assessment for TB among the participants was estimated to be 8.26 days with 95% CI (7.69 days - 8.84 days).

##create a stacked coloumn chart of number of patients by district and gender
```{r}
library(ggplot2)
ggplot(tb, aes(x=District, fill =Gender)) + geom_bar() + labs(little = "Number of TB patients by district and gender", x = "District", y="Number of patients", fill = "Gender") + theme_minimal()
```

##comment on whether age follows a normal distribution or not.
```{r}
hist(tb$Age)
shapiro.test(tb$Age)
mean(tb$Age)
round(mean(tb$Age),2)
median(tb$Age)
```
The histogram is skewed to the right. Shapiro test showed p <0.05
mean = 32.27 and median= 28
The variable age showed a non- normal distribution. 

##Stratify the number of abnormal parameters as low risk (1), moderate risk (2-3) and high risk (4 and greater than 4)
```{r}
tb <- tb |> mutate(risk = case_when(
  `No of Abnormal parameters`== 0 ~"No abnormal parameter",
  `No of Abnormal parameters`== 1 ~ "Low risk",
  `No of Abnormal parameters` <=3 ~ "Moderate risk",
  `No of Abnormal parameters`>= 4 ~ "High risk"
  )) |> 
  mutate (risk = factor(risk, levels = c("No abnormal parameter","Low risk", "Moderate risk","High risk")))
table(tb$risk)
sum(is.na(tb$risk))
round(prop.table(table(tb$risk))*100,2)
```
##Cross-Tabulate and find out the number of persons stratified by Medical Officer as being in low-risk despite the number of abnormal parameters being in high risk. 
```{r}
tb <- tb |> 
  mutate(mo_risk = case_when(
    grepl("^Low",`Final risk stratification BY THE MEDICAL OFFICER`) ~ "MO: Low",
    grepl("^Moderate",`Final risk stratification BY THE MEDICAL OFFICER`)~ "MO: Moderate",
    grepl("^High",`Final risk stratification BY THE MEDICAL OFFICER`)~ "MO: High"
  ))  |> 
    mutate(mo_risk = factor(mo_risk, levels = c("MO: Low","MO: Moderate","MO: High")))
table(tb$risk, tb$mo_risk, useNA="ifany")
sum(tb$risk =="High risk" & tb$mo_risk == "MO: Low", na.rm = TRUE)
```
The number of persons stratified by Medical Officer as being in low-risk despite the number of abnormal parameters being in high risk was 185.

##Create a descending order column chart showing number of persons having abnormal respiratory rate, bmi, haemoglobin and oxygen saturation
```{r}
abn <- tb |> 
  summarise(`Respiratory rate` =sum(`Respiratory rate`==1),
    BMI = sum(BMI ==1),
    Haemoglobin =sum(Haemoglobin == 1),
    `Oxygen saturation`= sum(`Oxygen saturation` == 1)) |> pivot_longer(everything(), names_to = "parameter", values_to = "n")

ggplot(abn, aes(x= reorder(parameter, -n), y=n)) + geom_col(fill = "seagreen") + geom_text(aes(label=n), vjust = -0.4)+ labs(title ="Patients with abnormal clinical parameters", x = "Parameter", y = "Number of patients") + theme_minimal()
```
##Create a subset of data set who are males and do not have blood in sputum

```{r}
Males_no_blood_sputum <- tb |> filter(Gender == "Male", `Blood in sputum`==0)
nrow(Males_no_blood_sputum)
```
There are 269 male participants who donot have blood in sputum.

##Anything else you want to report in this dataset. 

```{r}
tb <- tb |> 
  rename(no_abnormal_para =`No of Abnormal parameters`) |> 
  mutate(
    Gender = factor(Gender),
    District = factor(District),
    age_group = cut(Age,
                    breaks = c(0,14,29,44,59, Inf),
                    labels = c("0-14", "15-29", "30-44", "45-59", "60+"))
  )
table(tb$no_abnormal_para, tb$age_group)
```

##frequency table with percentages
```{r}
freq <- function(x) {
  t <- table(x, useNA = "ifany")
  data.frame(category = names(t),
  n = as.vector(t),
  percent = round(100* as.vector(prop.table(t)),1))
}
```
##Coulmns coded 1 = present/abnormal, 0= absent/normal

```{r}
symptoms <- c("Cough" = "Current symptoms /Cough",
    "Persistent fever" = "Current symptoms /Persistent fever",
    "Breathlessness" = "Current symptoms /Breathlessness",
    "Loss of weight" = "Current symptoms /Loss of weight",
    "Loss of appetite" = "Current symptoms /Loss of appetite",
    "Evening rise of temperature" = "Current symptoms /Evening rise of temperature",
    "Blood in sputum" = "Blood in sputum")
params <- c("BMI" = "BMI",
            "Haemoglobin" = "Haemoglobin",
            "SpO2" =  "SpO2",
            "RBS" = "RBS",
            "HIV" = "HIV",
            "Chest X-ray" = "Chest X-ray")
```
Distribution of study population as per Age, Gender and Location

```{r}
summary(tb$Age)
freq(tb$age_group)
freq(tb$Gender)
freq(tb$District)
```
##Distribution of Symptomatology
```{r}
sym_tab <- data.frame(
  symptom = names(symptoms),
  count = sapply (symptoms, function(v) sum(tb[[v]]==1)), 
  row.names =NULL
) |> 
  mutate(percent = round(100* count/ nrow(tb), 1)) |> 
  arrange(desc(count))
sym_tab
```
##Bivariate analysis of Delay in assessment and number of abnormal parameters by gender and district
```{r}
tb |> group_by(Gender) |> 
  summarise(n=n(),
            median_delay = median(delay), iqr_delay = IQR(delay),
            median_abn = median(no_abnormal_para), iqr_abn = IQR(no_abnormal_para)) 
tb |> group_by(Gender) |> 
  summarise(n=n(),
            median_delay = median(delay), iqr_delay = IQR(delay),
            median_abn = median(no_abnormal_para), iqr_abn = IQR(no_abnormal_para))
tb |> group_by(District) |> 
  summarise(n=n(),
            median_delay = median(delay), iqr_delay = IQR(delay),
            median_abn = median(no_abnormal_para), iqr_abn = IQR(no_abnormal_para))
wilcox.test(delay~Gender, data = tb)
wilcox.test(no_abnormal_para~Gender, data = tb)
kruskal.test(delay~District, data = tb)
kruskal.test(no_abnormal_para~District, data = tb)
ggplot(tb, aes(x= District, y= delay)) +
  geom_boxplot(fill ="steelblue", alpha =0.6) +
  labs(title = "Distribution as per Delay in assessment and district", y="Delay(days)") + theme_minimal()
```
